Radiocarbon measurement of CO~2~ from carbonate minerals using a gas ion source with an open split interface

### 2022-03-23 doc/AMS-15/AMS-15_paper/AMS-15_paper.Rmd
Add figures. Update blanks section. Other edits.

@@

![Inlet layout. Helium displacement gas is delivered to vial through a double needle, mixed sample and helium flows to an open split, and a glass capillary delivers gas to the ion source.](images/HGIS_open_split_drawing.svg)

@@

Reference: gilson

@@

### Measurement

@@

### Data reduction

@@

## System performance

Best currents for the source are around 10uA 12C- or roughly 1/10 of graphite currents. This is generally comparable with other hybrid sources REFS. Typical currents for this study were between 4-10uA???. The source typically produces best current at around 5uLs-1 CO2. Literature values give 1.5uLs-1 as optimal, but we've never seen that, and pressures, capillaries and flows in at least one of these papers doesn't make sense.

@@

Measured value of tank modern gas normalized to solid graphite standards ranges from agreeing with solid OX-I to 7% lower than expected for low current samples. Normalizing to tank CO2 is probably the best method to account for variability in hybrid source performance.

@@

Figure4: sample size vs current vs precision and accuracy

@@

To assess blank contribution and ratio dependence on sample current, vials of live and dead standard gas were measured as the displacement helium flow dilluted the samples, producing diminishing current over time. We used these data to develop a current-balance model following methods used to assess mass-dependent or constant contamination blanks in graphite targets (REFS). The constant contamination blank was also assessed directly by measuring pure helium in vials using the same method. Results from both tests are shown in figure 2. The inverse of the sample current is used to linearize the data, allowing assessment with a simple linear model. Fits from modern and dead dilution tests should trend towards the fraction modern of the blank, crossing at the inverse of blank current. Contaminant current measured directly agrees well with the model assessment, but has a higher fraction modern. This is likely due to a combination of low measurement precision for the direct measurement and high sensitivity of the model to variability in the measured fraction modern and current in the depletion samples (REF). This constant contamination blank is consistent with the fact that hybrid cathodes produce a minimum current with or without gas flow to the source. Given that this minimum current ranges from 100 to 400, this current likely makes up a large fraction of the method blank.

@@

Figure2: current dependent blank

@@

While dillution and direct measurement data show evidence for a constant blank with a variable effect depending on sample size, the blank correction may be simplified if blanks, standards, and unknown samples are the same size. If the fraction modern of the blank is 0.8 modern as estimated above, blank samples will show the most effect of a constant contamination blank with current. Figure 3 shows the fraction modern of measured blanks compared to their mean 12C- ion current. Fraction modern of measured blanks does decrease with increasing current, but the effect is small, less than Fm 0.01 over the range of currents measured. This change with current is roughly equal to the variablility between blanks, so the error due to the use of a current-dependent blank based on samples measured over multiple wheels will likely be larger than a current independent blank correction based on blanks measured on each wheel.

@@

```{r}
# Fix to use current form of processed data
carb_data <- map_dfr(list.files(here("data_analysed"), full.names = TRUE), read_csv) %>% 
  filter(!is.na(rec_num),
         Cur > 2) %>% 
  mutate(fm_diff = mean - fm_consensus,
         fm_sigma = sigma(mean, fm_consensus, merr),
         fmbc_diff = fm_lbc - fm_consensus,
         fmbc_sigma = sigma(fm_lbc, fm_consensus, merr))
```

Mean Fm of blanks: `r carb_data %>% filter(fm_consensus < 0.1 & Cur > 2) %>% pull(mean) %>% mean() %>% format(digits = 3)`

@@

@@

Fig. 1. Schematic of the open split inlet. Two valco valves provide the ability to admit helium and live or dead tank standards. The double needle  is housed in a VICI cross and allows displacement helium to flow into the sample vial and the sample-helium mixture to flow to the open split, where a capillary delivers the mixture to the ion source.

Fig. 2. Performance of the NOSAMS MCSNICS hybrid gas ion source. a. Modelled flows of helium and CO2 to the ion source during displacement of a pure CO2 vial sample with helium. b. Measured 12C- ion current during the same dillution experiment. c. Measured ion current for modelled CO2 flow. The stability of the source over a range of flow produces stable current for up to 40 m of measurement time.

Fig. 3. Relationship of sample size to C- ion current. There is no strong relationship between size and current for samples larger than 800 ug C.

Fig. 4. Assessment of blanks. (a) Current dependent blank model and data. Fm and current of measurements from dillution tests of live and dead gas are used to generate linear fits which should cross at the Fm and current of the constant current blank. Direct determination of Fm and current of the blank are also shown. (b) Fraction modern of large, radiocarbon dead samples compared to sample current. Blank rises with decreasing current, as expected, but may be treated as a single blank with a simple correction for full-size samples producing greater than 3 uA C- current.

Fig 5. Agreement of standards with graphite consensus values. Difference of HGIS measurements from consensus for individual samples are plotted against their expected value. Expected values are spread to avoid overlapping points. Error bars represent one sigma sample errors. The black line is the mean difference from expected value for all measurements, and the blue lines are one standard deviation of all measurements.

Fig 6. Comparison of Haiti core samples measured as graphite. a. Difference between HGIS and graphite measurements of the same materials vs. core depth (should this be mean of values?). Samples are offset on the horizontal axis for clarity. Error bars are measurement error for each sample. b) comparison of age-depth models produced using HGIS and graphite measurements.

>>>>

```{r, message = FALSE}
library(tidyverse)
library(hgis)
library(here)

carb <- readRDS(here("data_analysed/carb_all.rds")) %>% 
  map_dfr(2)
```

@@

@@

@@

## Measurement

@@

## Data reduction

@@

## Source performance

```{r}
carbcur <- carb %>%  
  map_dfr(2) %>% 
  compare_consensus() %>% 
  mutate(c12ua = he12C * 1E6)
samples <- read_csv(here("data/carb_data.csv")) %>% 
  select(wheel, pos, mass, he_added)
currents <- left_join(carbcur, samples, by = c("wheel", "pos")) %>% 
  filter(wheel != "USAMS020521")

curmean <- mean(currents$c12ua) %>% format(digits = 2) %>% paste("uA")
cursd <- sd(currents$c12ua)
```

Best currents for the source are around 10uA 12C- or roughly 1/10 of graphite currents. This is generally comparable with other hybrid sources REFS. Typical currents for this study were between 4-9uA and average current for all samples was `r curmean`. The source typically produces best current at around 5uLs-1 CO2. Literature values give 1.5uLs-1 as optimal, but we've never seen that, and pressures, capillaries and flows in at least one of these papers doesn't make sense.

@@

@@

@@

Measurement of radiocarbon-free reference materials shows that the average normalized ratio for these samples is XXX +- YYY (N=XXX). We investigated the sources of this method blank as a current-dependent blank and as a current-independent blank.

To assess blank contribution and ratio dependence on sample current, vials of live and dead standard gas were measured as the displacement helium flow diluted the samples, producing diminishing current over time. We used these data to develop a current-balance model following methods used to assess mass-dependent or constant contamination blanks in graphite targets (REFS). The constant contamination blank was also assessed directly by measuring pure helium in vials using the same method. Results from both tests are shown in figure 2. The inverse of the sample current is used to linearize the data, allowing assessment with a simple linear model. Fits from modern and dead dilution tests should trend towards the fraction modern of the blank, crossing at the inverse of blank current. Based on this assessment, we estimate the ratio of the blank as XXX pMC with a current of XXX nA.

The directly blank current (XXX nA) agrees well with the model assessment, but has a higher fraction modern (XXX pMC). This is likely due to a combination of low measurement precision for the direct measurement and high sensitivity of the model to variability in the measured fraction modern and current in the depletion samples (REF). This constant contamination blank is consistent with the fact that hybrid cathodes produce a minimum current with or without gas flow to the source. Given that this minimum current ranges from 100 nA to 400 nA, this current likely makes up a large fraction of the method blank.

@@

@@

While dilution and direct measurement data show evidence for a constant blank with a variable effect depending on sample size, the blank correction may be simplified if blanks, standards, and unknown samples are the same size. If the fraction modern of the blank is 0.8 modern as estimated above, blank samples will show the most effect of a constant contamination blank with current. Figure 3 shows the fraction modern of measured blanks compared to their mean 12C- ion current. Fraction modern of measured blanks does decrease with increasing current, but the effect is small, less than Fm 0.01 over the range of currents measured. This change with current is roughly equal to the variability between blanks, so the error due to the use of a current-dependent blank based on samples measured over multiple wheels will likely be larger than a current independent blank correction based on blanks measured on each wheel.

@@

```{r}
# Fix to use current form of processed data
blanks <- carb %>% 
  filter(rec_num %in% c(2138, 83028),
         he12C > 2E-6,
         sample_type == "B")
```

Mean Fm of blanks: `r mean(blanks$norm_ratio) %>% format(digits = 3)`
SD of Fm of blanks: `r sd(blanks$norm_ratio) %>% format(digits = 3)`

@@

# Figures

![Figure 1. Inlet layout. Helium displacement gas is delivered to vial through a double needle, mixed sample and helium flows to an open split, and a glass capillary delivers gas to the ion source.](images/HGIS_open_split_drawing.svg)

![Figure 2](figures/fig1a_flow_current_time.svg)

![Figure 2b](figures/fig1b_current_flow.svg)

![Figure 4](figures/fig2_current_dependent_blank.svg)

![Figure 4b](figures/fig3_cur_indep_blank.svg)

![Figure 5](figures/fig5_consensus_samples.svg)

![Figure 6](figures/fig6_core_carbonates.svg)

@@

Figure 1. Schematic of the open split inlet. Two valco valves provide the ability to admit helium and live or dead tank standards. The double needle is housed in a VICI cross and allows displacement helium to flow into the sample vial and the sample-helium mixture to flow to the open split, where a capillary delivers the mixture to the ion source.

Figure 2. Performance of the NOSAMS MCSNICS hybrid gas ion source. a. Modelled flows of helium and CO2 to the ion source during displacement of a pure CO2 vial sample with helium. b. Measured 12C- ion current during the same dillution experiment. c. Measured ion current for modelled CO2 flow. The stability of the source over a range of flow produces stable current for up to 40 m of measurement time.

Will probably skip this figure:
Figure 3. Relationship of sample size to C- ion current. There is no strong relationship between size and current for samples larger than 800 ug C.

Figure 4. Assessment of blanks. (a) Current dependent blank model and data. Fm and current of measurements from dillution tests of live and dead gas are used to generate linear fits which should cross at the Fm and current of the constant current blank. Direct determination of Fm and current of the blank are also shown. (b) Fraction modern of large, radiocarbon dead samples compared to sample current. Blank rises with decreasing current, as expected, but may be treated as a single blank with a simple correction for full-size samples producing greater than 3 uA C- current.

Figure 5. Agreement of standards with graphite consensus values. Difference of HGIS measurements from consensus for individual samples are plotted against their expected value. Expected values are spread to avoid overlapping points. Error bars represent one sigma sample errors. The black line is the mean difference from expected value for all measurements, and the blue lines are one standard deviation of all measurements.

Figure 6. Comparison of Haiti core samples measured as graphite. a. Difference between HGIS and graphite measurements of the same materials vs. core depth (should this be mean of values?). Samples are offset on the horizontal axis for clarity. Error bars are measurement error for each sample. b) comparison of age-depth models produced using HGIS and graphite measurements.
